Experiments and Modeling of the Autoignition of Methyl Pentanoate at Low to Intermediate Temperatures and Elevated Pressures in a Rapid Compression Machine

### 2017-09-26 revision-1/response/response.tex

\begin{response}
    Thank you for your comments. A detailed response is below.
\end{response}

\begin{reviewer}
    %
    I would like to ask the authors to check one item definitively before the article is published.
    I am concerned that at these higher pressures and equivalence ratios close to 1, there would be
    ignition which occurs during the compression stroke. This would not be noticed if some of the
    fuel condenses during the charging process and thereby you are dealing with lean mixtures than
    those cited in the paper. The authors are encouraged to make sure this is not the case by
    possibly sampling the charge after filling through a GC or by a quantitative absorption method
    as that used routinely by Stanford.
    %
\end{reviewer}

\begin{response}
    %
    With regards to ignition during the compression stroke, we have compared each reactive pressure
    trace to its corresponding non-reactive trace and found they are in good agreement from the
    beginning to the end of compression, indicating that there is no heat release during the
    compression stroke. With regards to the filling process, we have thoroughly checked the
    temperatures throughout the apparatus and found them to be quite consistent, including the
    mixing tanks, the piping connections, and the reaction chamber itself. Moreover, we have
    conducted several studies in the past utilizing GC to determine mole fractions of components.
    All of these studies determined that the vaporization procedure used on this apparatus is
    sufficient to produce a homogeneous mixture of fuel and oxidizer, even for fuels with relatively
    low vapor pressures. We have added references to these prior studies in the present manuscript
    near line 105:

    \begin{quote}
        %
        ``Previous work has shown this procedure to completely vaporize the fuel and prevent fuel
        cracking during the heating process~\cite{Weber2011a,Kumar2009,Das2012}.''
        %
    \end{quote}
    %
\end{response}

### 2017-09-26 revision-1/revised-paper/methyl-valerate.tex
Paper updates from Jackie

A recent review paper summarizes the work on methyl esters relevant to biodiesel
combustion~\cite{Coniglio2013}; the following summary focuses on ignition delay measurements, since
these are the focus of this paper. Autoignition of methyl butanoate (MB, \ce{C5H10O2}) has been
well-studied in both shock tube and rapid compression machine experiments~
\cite{Metcalfe2007,Walton2009,Dooley2008,Akih-Kumgeh2010,Akih-Kumgeh2011,Hadj-Ali2009,Kumar2016}.
The prevalence of MB data in the literature is largely due to the early identification of MB as a
potential surrogate fuel for biodiesel~\cite{Fisher2000}. However, the experiments have shown that
MB may not be an appropriate surrogate for biodiesel, due to its lack of negative temperature
coefficient (NTC) behavior, a requirement for a suitable biodiesel surrogate~\cite{Coniglio2013}.

@@

\begin{center}
    \captionsetup{type=figure}
    \includegraphics{figures/ignition_delay_definition.pdf}
    \caption{Definition of the ignition delays used in this work. The
    experiment in this figure is conducted for a \(\phi=2.0\) mixture with
    \mbox{\(\ce{Ar}/(\ce{N2}+\ce{Ar})=0.5\)}, \(P_0=\SI{0.7806}{\bar}\),
    \(T_0=\SI{373}{\K}\), \(P_C=\SI{14.92}{\bar}\), \(T_C=\SI{720}{\K}\),
    \(\tau=\SI{27.56\pm0.89}{\ms}\), and \(\tau_1=\SI{16.60\pm0.46}{\ms}\).}
    \label{fig:ign-delay-def}
\end{center}

@@

In addition to the reactive experiments, non-reactive experiments are conducted by replacing \ce{O2}
with \ce{N2} to determine the influence of machine-specific behavior on the experimental conditions
and permit the calculation of the EOC temperature via the isentropic relations between pressure and
temperature~\cite{Lee1998}. The EOC temperature is calculated by the procedure described in
\cref{sec:computational-methods}.

@@

Mixtures are prepared in stainless steel mixing tanks, \SI{17}{\L} and \SI{15}{\L} in size. The
proportions of reactants in the mixture are determined by specifying the absolute mass of the fuel,
the equivalence ratio, and the ratio of \(\ce{Ar}:\ce{N2}\) in the oxidizer. Mixtures are
made by first vacuuming the mixing tanks to an ultimate pressure less than \SI{5}{\torr}. Since MV
is a liquid with a relatively small vapor pressure at room temperature and pressure, it is measured
gravimetrically to within \SI{0.01}{\g} of the specified value. The fuel is injected into the mixing
tank through a septum. Proportions of \ce{O2}, \ce{Ar}, and \ce{N2} are added manometrically at room
temperature and the total pressure is measured by an Omega Engineering MMA100V10T2D0T4A6 type static
pressure transducer. The same transducer is used to measure the pressure of the reactants prior to
an experiment.

The RCM is equipped with heaters to control the initial temperature of the mixture. After filling in
the components to the mixing tanks, the heaters are switched on and the system is allowed
\SI{1.5}{\hour} to come to steady state. The mixing tanks are also equipped with magnetic stir bars
so the reactants are well mixed for the duration of the experiments. Previous work has shown this
procedure to completely vaporize the fuel and prevent fuel cracking during the heating process
\cite{Weber2011a,Kumar2009,Das2012}.

@@

The initial temperature is chosen such that the saturated vapor pressure \((P_{\text{sat}})\) of the
fuel at the initial temperature is at least twice the partial pressure of the fuel in the mixing
tank. The Antoine equation
%
\begin{equation}\label{eq:antoine}
    \logten{P_{\text{sat}}} = A - \frac{B}{T - C}
\end{equation}
%
is used to model the saturated vapor pressure of MV as a function of temperature \((T)\), where
\(A\), \(B\), and \(C\) are substance-specific coefficients, given in units of \si{\kelvin} and
\si{\kPa}. Coefficients for \cref{eq:antoine} are given in the literature by~\citet{Ortega2003},
\citet{Camacho2007}, and~\citet{Stephenson1987}. Unfortunately, the values of the coefficients are
different among all three references. Therefore, coefficients for use in \cref{eq:antoine} are
determined in this work by least squares fitting of the data of \citet{Ortega2003},
\citet{vanGenderen2002}, and \citet{Verevkin2008} using the \verb|curve_fit()| function of
SciPy~\cite{Jones2001} version 0.19.0. \Cref{fig:vapor-pressure} shows that the coefficients fitted
with this procedure give good agreement with the experimental data; values for the coefficients
computed in this work and in the literature works are given in \cref{tab:antoine}. The data and code
used to calculate the coefficients are provided in the Supplementary Material.

\begin{center}
    \captionsetup{type=table}
    \caption{Antoine Equation coefficients computed in this work and from the
    literature, in units of \si{\kelvin} and \si{\kPa}. The $2\sigma$ confidence interval is
    estimated by taking the square root of the diagonals of the covariance matrix returned from
    \texttt{curve\_fit()}}
    \label{tab:antoine}
    \begin{tabular}{cccccc}
        \toprule
        & $A$ & $B$ & $C$ & $T_{\text{min}}$, \si{\K} & $T_{\text{max}}$, \si{\K} \\
        \midrule
        This Work & 6.4030 & 1528.69 & 52.881 & 274.9 & 417.18 \\
        $2\sigma$ Confidence Interval & 0.0919 & 53.47 & 4.934 & --- & --- \\
        \citet{Ortega2003} & 6.23175 & 1429.00 & 62.30 & 364.75 & 417.18 \\
        \citet{Camacho2007} & 5.9644 & 1281.06 & 75.94 & 281 & 547 \\
        \citet{Stephenson1987} & 6.62646 & 1658.4 & 42.09 & 297 & 411 \\
        \bottomrule
    \end{tabular}
\end{center}

@@

To the best of our knowledge, there are three mechanisms for MV combustion available in the
literature. The first two, by \citet{Korobeinichev2015} and \citet{Dmitriev2015}, were developed to
simulate flames, and do not include the low-temperature chemistry necessary to simulate the
conditions in these experiments. The third model was developed by \citet{Dievart2013} and includes
low-temperature chemistry of MV, although it was only validated by comparison with flame extinction
limits. In converting this mechanism for use in Cantera, we found that there were many species in
the thermodynamic database with multiple data entries. For most of these species the thermodynamic
data is identical. However, some species are not exact duplicates. For these species, it is not
clear from the thermodynamic database file which data set should be preferred. Since Cantera (and
CHEMKIN) choose the first instance of a duplicate species to be used, we retained the first entry
for all duplicated species. The detailed model of \citet{Dievart2013} includes 1105 species and 7141
reactions, and the Cantera formatted input file is available in the Supplementary Material.

@@

In addition to using a mechanism from the literature, we investigate the use of an automatic
mechanism generator, the open-source Reaction Mechanism Generator (RMG)~\cite{Allen2012} version
2.1.0. The Python version of RMG is used, which requires Python 2.7, and version 2.1.0 of the RMG
database is used. The final RMG model contains 427 species and 13640 reactions. Note that the number
of species is much lower than the \citet{Dievart2013} model because the RMG model focuses on only
one fuel (MV), but the number of reactions is substantially higher. The input file used to generate
the model is available in the Supplementary Material. In addition, the CHEMKIN and Cantera formatted
input files for the RMG model are available in the Supplementary Material.

@@

\Cref{fig:ignition-delays} shows the ignition delay results measured in this study. Filled markers
denote the overall ignition delay and hollow markers indicate the first-stage ignition delay.
Vertical error bars are drawn on the symbols to represent the \(2\sigma\) uncertainty in the
ignition delay; for many of the experiments, the uncertainty is approximately the same size as the
data point, so the error bar is hidden. Horizontal error bars are shown on the first and last points
of each equivalence ratio indicating the estimated uncertainty in the EOC temperature of
$\pm\SI{1}{\percent}$~\cite{Weber2015}. \cref{fig:ignition-delays}a shows the results for a
compressed pressure of \SI{15}{\bar}, while \cref{fig:ignition-delays}b shows the results for a
compressed pressure of \SI{30}{\bar}. Note that $\phi=2.0$ results were not collected for
\SI{30}{\bar}, so there are no red triangle data points in \cref{fig:ignition-delays}b. A summary of
the ignition delay data is available as a comma-separated value file in the Supplementary Material.

@@

\citet{Hadj-Ali2009} also observed two-stage ignition of MV in stoichiometric mixtures, stating that
``[m]ethyl pentanoate... was more reactive [than methyl butanoate] with a limit below which
autoignition no longer occurs observed at \(T_c = \SI{670}{\K}\) and \(P_c = \SI{11.4}{\bar}\). At
this temperature, the autoignition occurred in two stages with a clearly identified cool flame
event.'' However, we do not find two stage ignition for the similar pressure of
\(P_C = \SI{15}{\bar}\) in this study. We note that the stated temperature of the experiment from
the work of \citet{Hadj-Ali2009} (\SI{670}{\K}) is much lower than the lowest temperature we
considered in this work at \SI{15}{\bar}, \(\phi=1.0\) (\SI{828}{\K}). We did not conduct
experiments at lower temperatures because the work of \citet{Mittal2006} showed that the temperature
field in the RCM reaction chamber was uniform for approximately \SI{100}{\ms} after the EOC, and our
measured ignition delay at \SI{15}{\bar}, \(\phi=1.0\), and \SI{828}{\K} is \SI{92.14}{\ms}.

However, we note NTC behavior of the ignition delay and two-stage ignition at the higher pressure of
\SI{30}{\bar}, and at higher temperatures than those reported for two-stage ignition in the study of
\citet{Hadj-Ali2009}. The trend of NTC behavior shifting to higher temperatures with increasing
pressure can be seen in other classes of fuels. \citet{Kukkadapu2012} found a similar trend in
gasoline composed of iso-alkanes, n-alkanes, cyclo-alkanes, aromatics, and olefins.
\citet{Kukkadapu2012} attributed the shift of the NTC region to the reactions between the
hydroperoxyalkyl radical (\ce{QOOH}) and \ce{O2} becoming more dominant than the unimolecular
decomposition of \ce{QOOH} at higher pressures. Similar trends could occur for the hydroperoxyalkyl
radicals of MV.

To further understand the effect of the methyl ester functional group on the NTC region of ignition
delay, we compare with the alkane and alcohol with 5-carbon alkyl chains, n-pentane and n-pentanol.
n-Pentane and MV have the same fuel mole percentage for stoichiometric mixtures in air
(\SI{2.56}{\percent}), while n-pentanol has a fuel mole percentage of \SI{2.72}{\percent} for
stoichiometric conditions. \citet{Ribaucour1998} and \citet{Bugler2015} found the NTC region for
n-pentane to be between \SIlist{760;910}{\K} at pressures near \SI{10}{\atm}. As we will compare
with our MV data at \SI{30}{\bar}, we note that increasing the pressure tends to shift the NTC to
higher temperatures, as mentioned previously \cite{Kukkadapu2012}. \citet{Heufer2013} found NTC
behavior for n-pentanol in the range of \SIrange{770}{900}{\K} at \SI{30}{\bar}. In this study, we
find the NTC window for MV at \SI{30}{\bar} to be between \SIlist{720;800}{\K}. Therefore, it
appears that the methyl ester functional group causes the NTC range to occur at lower temperature as
compared to alkanes and alcohols with similar alkyl chain lengths. This result was also noted by
\citet{Hadj-Ali2009} for methyl hexanoate as the fuel.

@@

\begin{center}
    \captionsetup{type=figure}
    \includegraphics[width=\textwidth]{figures/pressure-traces.pdf}
    \caption{Selected pressure traces around the NTC region of ignition delay.
    a) \(\phi=1.0,\ P_C=\SI{30}{\bar}\), b) \(\phi=2.0,\ P_C=\SI{15}{\bar}\).}
    \label{fig:pressure-traces}
\end{center}

@@

At \SI{15}{\bar}, the experimental ignition delays are under-predicted by the \citet{Dievart2013}
model for the three equivalence ratios shown. For the \(\phi =\) \numlist{0.25;0.5} conditions, the
model appears to be predicting an NTC region of the ignition delays as the temperature decreases, as
judged by the increasing curvature of the simulations, although such a trend is not observed in the
data. However, at \(\phi = 2.0\), the model does not predict the presence of an NTC region, although
one is present in the experiments. Nonetheless, the agreement seems to be improving as the
temperature is decreased. Comparing the \citet{Dievart2013} model to the stoichiometric data at
\SI{30}{\bar}, we find a similar trend as the \(\phi=2.0,\ P_C=\SI{15}{\bar}\) data. The model does
not predict the NTC region found experimentally for the \(\phi = 1.0\) experiments, but the
agreement improves as the temperature decreases. Interestingly, two-stage ignition is predicted for
all of the \(\phi=1.0\) and \(\phi=2.0\) data shown in \cref{fig:simulation-comparison}. However,
the first-stage ignition delays are \SIrange{0.1}{0.5}{\ms} less than the overall ignition delays,
and are not shown on \cref{fig:simulation-comparison} because they are nearly indistinguishable from
the overall ignition delay.

@@

To further understand the model of \citet{Dievart2013}, we have conducted adiabatic, constant volume
simulations (called CONV simulations), as these simulations are not linked to a particular
experiment by the volume trace and can be conducted over a wide range of temperatures.
\Cref{fig:conv-comparison} compares the CONV simulations to the RCM simulations shown in
\cref{fig:simulation-comparison}. From \cref{fig:conv-comparison}, it can be seen that the curvature
in the RCM simulations at $P_C = \SI{15}{\bar}$, $\phi =$ \numlist{0.25;0.5} is related to the NTC
region of the ignition delay, while the lack of curvature in the $P_C = \SI{15}{\bar}$, $\phi =
\num{2.0}$ and the $P_C = \SI{30}{\bar}$, $\phi = \num{1.0}$ simulations is because those lie on the
low-temperature side of the predicted NTC. It is clear from \cref{fig:conv-comparison} that the
model of \citet{Dievart2013} predicts the NTC to occur at too high a temperature relative to the
experiments.

To elucidate the underlying reasons for the disagreement between the \citet{Dievart2013} model and
the data, we constructed an additional model using RMG (see
\cref{sec:reaction-mechanism-generator}). As can be seen in \cref{fig:simulation-comparison}a, the
agreement between the RMG model (dashed lines) and the experimental data is similar to the
\citet{Dievart2013} model for the \SI{15}{\bar}, \(\phi =\) \numlist{0.25; 0.5} data. Moreover, the
RMG model predicts a similar NTC region as temperature is decreasing. For the \SI{15}{\bar}, \(\phi =
2.0\) data, the RMG model tends to over-predict the low-temperature overall ignition delays (i.e.,
those to the right of the experimental NTC region on the Arrhenius plot), and does not predict the
NTC region found experimentally. As before, the trend at \SI{30}{\bar}, \(\phi=1.0\) is similar to
the \SI{15}{\bar}, \(\phi=2.0\) data; the RMG model over-predicts the low-temperature ignition
delays and does not predict the experimental NTC region. Finally, as in the \citet{Dievart2013}
model, two-stage ignition is predicted for all of the \(\phi=1.0\) and \(\phi=2.0\) data shown in
\cref{fig:simulation-comparison}. However, the first-stage ignition delays are
\SIrange{0.1}{0.5}{\ms} less than the overall ignition delays, and are not shown on
\cref{fig:simulation-comparison} because they are nearly indistinguishable from the overall ignition
delay.

It is clear that neither model is able to predict the ignition delays of MV particularly well. In
addition to the poor agreement shown in \cref{fig:simulation-comparison}, the simulations for
\(P_C = \SI{15}{\bar}\), \(\phi=1.0\) and \(P_C = \SI{30}{\bar}\), \(\phi=\) \numlist{0.25;0.5;2.0}
showed substantial heat release during the compression stroke (i.e., the simulations are much too
reactive), and so these conditions aren't compared in \cref{fig:simulation-comparison}. We note
again that the model by \citet{Dievart2013} was validated for MV combustion only by comparison to
flame extinction limits, so the disagreement is not wholly surprising.

@@

The third source of error may lie in the estimation of the thermodynamic properties of the species,
particularly the fuel radicals. In the work of \citet{Dievart2013}, the program THERM
\cite{Ritter1991} is used to estimate thermodynamic values using the group additivity method. In the
RMG model constructed in this work, RMG itself estimates the thermodynamic properties of the
molecules also using the group additivity method. Nonetheless, the two models have differing
predictions of the thermodynamic properties of the species in the model, particularly the fuel and
its radicals. The values of the heats of formation of the fuel and its \ce{H}-atom abstraction
radicals are shown in \cref{tab:heat-of-formation}; the radicals are labeled according to the
convention shown in \cref{fig:mv-structure}.

>>>>

A recent review paper summarizes the work on methyl esters relevant to biodiesel
combustion~\cite{Coniglio2013}; the following summary focuses on ignition delay measurements, since
these are the focus of this paper. Autoignition of methyl butanoate (MB, \ce{C5H10O2}) has been
well-studied in both shock tube and rapid compression machine experiments~
\cite{Metcalfe2007,Walton2009,Dooley2008,Akih-Kumgeh2010,Akih-Kumgeh2011,Hadj-Ali2009,Kumar2016}.
The prevalence of MB data in the literature is largely due to the early identification of MB as a
potential surrogate fuel for biodiesel~\cite{Fisher2000}. However, the literature experiments have
shown that MB may not be an appropriate surrogate for biodiesel, due to its lack of negative
temperature coefficient (NTC) behavior, a requirement for a suitable biodiesel
surrogate~\cite{Coniglio2013}.

@@

\begin{center}
    \captionsetup{type=figure}
    \includegraphics{figures/ignition_delay_definition.pdf}
    \caption{Definition of the ignition delays used in this work. The experiment in this figure is
    conducted for a \(\phi=2.0\) mixture with \mbox{\(\ce{Ar}/(\ce{N2}+\ce{Ar})=0.5\)},
    \(P_0=\SI{0.7806}{\bar}\), \(T_0=\SI{373}{\K}\), \(P_C=\SI{14.92}{\bar}\), \(T_C=\SI{720}{\K}\),
    \(\tau=\SI{27.56\pm0.89}{\ms}\), and \(\tau_1=\SI{16.60\pm0.46}{\ms}\). The non-reacting
    pressure traces by replacing \ce{O2} with \ce{N2} is also shown for reference.}
    \label{fig:ign-delay-def}
\end{center}

@@

In addition to the reactive experiments, non-reactive experiments are conducted by replacing \ce{O2}
with \ce{N2} to determine the influence of machine-specific behavior on the experimental conditions
(see \cref{fig:ign-delay-def}) and permit the calculation of the EOC temperature via the isentropic
relations between pressure and temperature~\cite{Lee1998}. The EOC temperature is calculated by the
procedure described in \cref{sec:computational-methods}.

@@

Mixtures are prepared in stainless steel mixing tanks, \SI{17}{\L} and \SI{15}{\L} in size. The
proportions of reactants in the mixture are determined by specifying the absolute mass of the fuel,
the equivalence ratio, and the ratio of \(\ce{Ar}:\ce{N2}\) in the oxidizer. Mixtures are made by
first vacuuming the mixing tanks to an ultimate pressure less than \SI{5}{\torr}. Since MV is a
liquid with a relatively small vapor pressure at room temperature and pressure, it is measured
gravimetrically to within \SI{0.01}{\g} of the specified value. The fuel is injected into the mixing
tank through a septum. Proportions of \ce{O2}, \ce{Ar}, and \ce{N2} are added manometrically at room
temperature and the total pressure is measured by an Omega Engineering MMA100V10T2D0T4A6 type static
pressure transducer. The same transducer is used to measure the pressure of the reactant mixture
prior to an experiment.

The RCM is equipped with heaters to control the initial temperature of the mixture. After filling in
the components to the mixing tanks, the heaters are switched on and the system is allowed
\SI{1.5}{\hour} to come to steady state. The mixing tanks are also equipped with magnetic stir bars
so the reactants are well mixed for the duration of the experiments. Previous work has shown this
procedure to completely vaporize the fuel and prevent fuel cracking during the heating
process~\cite{Weber2011a,Kumar2009,Das2012}.

The initial temperature is chosen such that the saturated vapor pressure \((P_{\text{sat}})\) of the
fuel at the initial temperature is at least twice the partial pressure of the fuel in the mixing
tank. The Antoine equation
%
\begin{equation}\label{eq:antoine}
    \logten{P_{\text{sat}}} = A - \frac{B}{T - C}
\end{equation}
%
is used to model the saturated vapor pressure of MV as a function of temperature \((T)\), where
\(A\), \(B\), and \(C\) are substance-specific coefficients, given in units of \si{\kelvin} and
\si{\kPa}. Coefficients for \cref{eq:antoine} are given in the literature by~\citet{Ortega2003},
\citet{Camacho2007}, and~\citet{Stephenson1987}. Unfortunately, the values of the coefficients are
different among all three references. Therefore, coefficients for use in \cref{eq:antoine} are
determined in this work by least squares fitting of the data of \citet{Ortega2003},
\citet{vanGenderen2002}, and \citet{Verevkin2008} using the \verb|curve_fit()| function of
SciPy~\cite{Jones2001} version 0.19.0. \Cref{fig:vapor-pressure} shows that the coefficients fitted
with this procedure give good agreement with the experimental data; values for the coefficients
computed in this work and reported in the literature works are given in \cref{tab:antoine}. The data
and code used to calculate the coefficients are provided in the Supplementary Material.

@@

\begin{center}
    \captionsetup{type=table}
    \caption{Antoine Equation coefficients computed in this work and obtained from the literature,
    in units of \si{\kelvin} and \si{\kPa}. The $2\sigma$ confidence interval is estimated by taking
    the square root of the diagonals of the covariance matrix returned from \texttt{curve\_fit()}}
    \label{tab:antoine}
    \begin{tabular}{cccccc}
        \toprule
        & $A$ & $B$ & $C$ & $T_{\text{min}}$, \si{\K} & $T_{\text{max}}$, \si{\K} \\
        \midrule
        This Work & 6.4030 & 1528.69 & 52.881 & 274.9 & 417.18 \\
        $2\sigma$ Confidence Interval & 0.0919 & 53.47 & 4.934 & --- & --- \\
        \citet{Ortega2003} & 6.23175 & 1429.00 & 62.30 & 364.75 & 417.18 \\
        \citet{Camacho2007} & 5.9644 & 1281.06 & 75.94 & 281 & 547 \\
        \citet{Stephenson1987} & 6.62646 & 1658.4 & 42.09 & 297 & 411 \\
        \bottomrule
    \end{tabular}
\end{center}

@@

To the best of our knowledge, there are three mechanisms for MV combustion available in the
literature. The first two, by \citet{Korobeinichev2015} and \citet{Dmitriev2015}, were developed to
simulate flames, and do not include the low-temperature chemistry necessary to simulate the
conditions in the current RCM experiments. The third model was developed by \citet{Dievart2013} and
includes low-temperature chemistry of MV, although it was only validated by comparison with flame
extinction limits. In converting this mechanism for use in Cantera, we found that there were many
species in the thermodynamic database with multiple data entries. For most of these species the
thermodynamic data is identical. However, some species are not exact duplicates. For these species,
it is not clear from the thermodynamic database file which data set should be preferred. Since
Cantera (and CHEMKIN) choose the first instance of a duplicate species to be used, we retained the
first entry for all duplicated species. The detailed model of \citet{Dievart2013} includes 1105
species and 7141 reactions, and the Cantera formatted input file is available in the Supplementary
Material.

@@

In addition to using a mechanism from the literature, we investigate the use of an automatic
mechanism generator, the open-source Reaction Mechanism Generator (RMG)~\cite{Allen2012} version
2.1.0. The Python version of RMG is used, which requires Python 2.7, and version 2.1.0 of the RMG
database is used. The final RMG model contains 427 species and 13640 reactions. Note that the number
of species is much lower than the \citet{Dievart2013} model because the RMG model focuses on only
one fuel (MV), but the number of reactions is substantially higher. The input file used to generate
the model is available in the Supplementary Material. In addition, the CHEMKIN and Cantera formatted
input files for the RMG-generated model are available in the Supplementary Material.

@@

\Cref{fig:ignition-delays} shows the ignition delay results measured in this study. Filled markers
denote the overall ignition delay and hollow markers indicate the first-stage ignition delay.
Vertical error bars are drawn on the symbols to represent the \(2\sigma\) uncertainty in the
ignition delay; for many of the experiments, the uncertainty is approximately the same size as the
data point, so the error bar is hidden. Horizontal error bars are shown on the first and last points
of each equivalence ratio indicating the estimated uncertainty in the EOC temperature of
$\pm\SI{1}{\percent}$~\cite{Weber2015}. \cref{fig:ignition-delays}a shows the results for a
compressed pressure of \SI{15}{\bar}, while \cref{fig:ignition-delays}b shows the results for a
compressed pressure of \SI{30}{\bar}. Note that $\phi=2.0$ results were not collected for
\SI{30}{\bar}, so there are no red triangle data points in \cref{fig:ignition-delays}b. A summary of
the ignition delay data is available as a comma-separated value file in the Supplementary Material,
including the mixture conditions for each experiment, the initial conditions, the compressed
conditions, and the ignition delays and their associated errors.

@@

\citet{Hadj-Ali2009} also observed two-stage ignition of MV in stoichiometric mixtures, stating that
``[m]ethyl pentanoate... was more reactive [than methyl butanoate] with a limit below which
autoignition no longer occurs observed at \(T_c = \SI{670}{\K}\) and \(P_c = \SI{11.4}{\bar}\). At
this temperature, the autoignition occurred in two stages with a clearly identified cool flame
event.'' However, we do not find two-stage ignition for the similar pressure of
\(P_C = \SI{15}{\bar}\) in this study. We note that the stated temperature of the experiment from
the work of \citet{Hadj-Ali2009} (\SI{670}{\K}) is much lower than the lowest temperature we
considered in this work at \SI{15}{\bar}, \(\phi=1.0\) (\SI{828}{\K}). We did not conduct
experiments at lower temperatures because the work of \citet{Mittal2006} showed that the temperature
field in the RCM reaction chamber was uniform for approximately \SI{100}{\ms} after the EOC, and our
measured ignition delay at \SI{15}{\bar}, \(\phi=1.0\), and \SI{828}{\K} is \SI{92.14}{\ms}.

However, we find NTC behavior of the overall ignition delay and two-stage ignition at the higher
pressure of \SI{30}{\bar}, and at higher temperatures than those reported for two-stage ignition in
the study of \citet{Hadj-Ali2009}. The trend of NTC behavior shifting to higher temperatures with
increasing pressure can be seen in other classes of fuels. \citet{Kukkadapu2012} found a similar
trend in gasoline composed of iso-alkanes, n-alkanes, cyclo-alkanes, aromatics, and olefins.
\citet{Kukkadapu2012} attributed the shift of the NTC region to the reactions between the
hydroperoxyalkyl radical (\ce{QOOH}) and \ce{O2} becoming more dominant than the unimolecular
decomposition of \ce{QOOH} at higher pressures. Similar trends could occur for the hydroperoxyalkyl
radicals of MV.

To further understand the effect of the methyl ester functional group on the NTC region of ignition
delay, we compare with the alkane and alcohol with 5-carbon alkyl chains, n-pentane and n-pentanol.
n-Pentane and MV have the same fuel mole percentage for stoichiometric mixtures in air
(\SI{2.56}{\percent}), while n-pentanol has a fuel mole percentage of \SI{2.72}{\percent} for
stoichiometric conditions. \citet{Ribaucour1998} and \citet{Bugler2015} found the NTC region for
n-pentane to be between \SIlist{760;910}{\K} at pressures near \SI{10}{\atm}. As we will compare
with our MV data at \SI{30}{\bar}, we note that increasing the pressure tends to shift the NTC to
higher temperatures, as mentioned previously~\cite{Kukkadapu2012}. \citet{Heufer2013} found NTC
behavior for n-pentanol in the range of \SIrange{770}{900}{\K} at \SI{30}{\bar}. In this study, we
find the NTC window for MV at \SI{30}{\bar} to be between \SIlist{720;800}{\K}. Therefore, it
appears that the methyl ester functional group causes the NTC range to occur at lower temperature as
compared to alkanes and alcohols with similar alkyl chain lengths. This result was also noted by
\citet{Hadj-Ali2009} for methyl hexanoate as the fuel.

@@

\begin{center}
    \captionsetup{type=figure}
    \includegraphics[width=\textwidth]{figures/pressure-traces.pdf}
    \caption{Selected pressure traces around the NTC region of ignition delay.  a) \(\phi=1.0\),
    \(P_C=\SI{30}{\bar}\), b) \(\phi=2.0\), \(P_C=\SI{15}{\bar}\). The corresponding non-reactive
    pressure traces are also included for reference.}
    \label{fig:pressure-traces}
\end{center}

@@

At \SI{15}{\bar}, the experimental overall ignition delays are under-predicted by the
\citet{Dievart2013} model for the three equivalence ratios shown. For the \(\phi =\)
\numlist{0.25;0.5} conditions, the model appears to be predicting an NTC region of the overall
ignition delays as the temperature decreases, as judged by the increasing curvature of the
simulations, although such a trend is not observed apparent in the figure. However, at \(\phi =
2.0\), the model does not predict the presence of an NTC region, although one is present in the
experiments. Nonetheless, the agreement seems to be improving as the temperature is decreased.
Comparing the \citet{Dievart2013} model to the stoichiometric data at \SI{30}{\bar}, we find a
similar trend as the \(\phi=2.0\), \(P_C=\SI{15}{\bar}\) data. The model does not predict the NTC
region found experimentally for the \(\phi = 1.0\), \(P_C=\SI{30}{\bar}\) experiments, but the
agreement improves as the temperature decreases. Interestingly, two-stage ignition is predicted for
all of the \(\phi=1.0\) and \(\phi=2.0\) data shown in \cref{fig:simulation-comparison}. However,
the first-stage ignition delays are \SIrange{0.1}{0.5}{\ms} less than the overall ignition delays,
and are not shown on \cref{fig:simulation-comparison} because they are nearly indistinguishable from
the overall ignition delay.

@@

To further understand the model of \citet{Dievart2013}, we have conducted adiabatic, constant volume
simulations (called CONV simulations), as these simulations are not linked to a particular
experiment by the volume trace and can be conducted over a wide range of temperatures. In the CONV
simulations, overall ignition delay is defined as an increase in the temperature of \SI{400}{\K}
over the initial temperature. \Cref{fig:conv-comparison} compares the CONV simulations to the RCM
simulations shown in \cref{fig:simulation-comparison}. From \cref{fig:conv-comparison}, it can be
seen that the curvature in the RCM simulations at \(P_C = \SI{15}{\bar}\), \(\phi =\)
\numlist{0.25;0.5} is related to the NTC region of the overall ignition delay, while the lack of
curvature in the \(P_C = \SI{15}{\bar}\), \(\phi = \num{2.0}\) and the \(P_C = \SI{30}{\bar}\),
\(\phi = \num{1.0}\) simulations is because those lie on the low-temperature side of the predicted
NTC. It is clear from \cref{fig:conv-comparison} that the model of \citet{Dievart2013} predicts the
NTC to occur at too high a temperature relative to the experiments.

To elucidate the underlying reasons for the disagreement between the \citet{Dievart2013} model and
the data, we conduct simulations with an additional model constructed using RMG (see
\cref{sec:reaction-mechanism-generator}). As can be seen in \cref{fig:simulation-comparison}a, the
agreement between the RMG model (dashed lines) and the experimental data is similar to the
\citet{Dievart2013} model for the \(P_C = \SI{15}{\bar}\), \(\phi =\) \numlist{0.25; 0.5} data.
Moreover, the RMG model predicts a similar NTC region as temperature is decreasing. For the \(P_C =
\SI{15}{\bar}\), \(\phi = 2.0\) data, the RMG model tends to over-predict the low-temperature
overall ignition delays (i.e., those to the right of the experimental NTC region on the Arrhenius
plot), and does not predict the NTC region found experimentally. As before, the trend at \(P_C =
\SI{30}{\bar}\), \(\phi=1.0\) is similar to the \(P_C = \SI{15}{\bar}\), \(\phi=2.0\) data; the RMG
model over-predicts the low-temperature overall ignition delays and does not predict the
experimental NTC region. Finally, as in the \citet{Dievart2013} model, two-stage ignition is
predicted for all of the \(\phi=1.0\) and \(\phi=2.0\) data shown in
\cref{fig:simulation-comparison}. However, the first-stage ignition delays are
\SIrange{0.1}{0.5}{\ms} less than the overall ignition delays, and are not shown on
\cref{fig:simulation-comparison} because they are nearly indistinguishable from the overall ignition
delay.

It is clear that neither model is able to predict the ignition delays of MV particularly well. In
addition to the poor agreement shown in \cref{fig:simulation-comparison}, the simulations for
\(P_C = \SI{15}{\bar}\), \(\phi=1.0\) and \(P_C = \SI{30}{\bar}\), \(\phi=\) \numlist{0.25;0.5;2.0}
showed substantial heat release during the compression stroke (i.e., the simulations are much too
reactive), and so these conditions are not compared in \cref{fig:simulation-comparison}. We note
again that the model by \citet{Dievart2013} was validated for MV combustion only by comparison to
flame extinction limits, so the disagreement is not wholly surprising.

@@

The third source of error may lie in the estimation of the thermodynamic properties of the species,
particularly the fuel radicals. In the work of \citet{Dievart2013}, the program
THERM~\cite{Ritter1991} was used to estimate thermodynamic values using the group additivity method.
In the RMG model constructed in this work, RMG itself estimates the thermodynamic properties of the
molecules also using the group additivity method. Nonetheless, the two models have differing
predictions of the thermodynamic properties of the species in the model, particularly the fuel and
its radicals. The values of the heats of formation of the fuel and its \ce{H}-atom abstraction
radicals are shown in \cref{tab:heat-of-formation}; the radicals are labeled according to the
convention shown in \cref{fig:mv-structure}.
